---
title: "Chapter 2 Exercises"
author: "Aidan O'Keeffe"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 1
### Problem
For each of parts (a) through (d), indicate whether we would generally expect the performance of a flexible statistical learning method to be better or worse than an inflexible model. Justify your answer.

a) The sample size $n$ is extremely large, and the number of predictors $p$ is small.
b) The number of predictors $p$ is extremely large, and the number of observations $n$ is small.
c) The relationship between the predictors and response is highly non-linear.
d) The variance of the error terms, i.e. $\sigma^2 = \mbox{Var}(\epsilon)$, is extremely high.

### Solution
a) A flexible model would likely perform much better than an inflexible model for two reasons. First, the relationship between the predictors and the response becomes more and more clear as the sample size grows, and since such relationships are almost always nonlinear, a flexible method could capitalize on the abundance of data to find such relationships. Second, because the number of predictors is small, we are in a low dimensional setting, meaning that we do not suffer from the curse of dimensionality, hence we could even expect extremely flexible nonparametric models to show good performance.
b) In this case, we would expect better performance from a less flexible model. Given a small $n$, as $p$ increases, overfitting become inevitable, which would lead a high test error rate. A less flexible model could identify a broad trend from what data _is_ available while not fitting the sample too closely.
c) If the relationship between the predictors and the response is highly non-linear, then a flexible method will perform better because it will provide a smaller model bias than a less flexible method.
d) The problem with high variance errors is that they obscure the true response-predictor relationship. In this case, a flexible method will be fit to the errors, almost certainly leading to poor test performance. As with scenario (b), we'd be better off using a less flexible model and identifying a broad trend in the data, thereby not fitting the noise too closely.

## Exercise 2
### Problem
Explain whether each scenario is a classification or regression problem, and indicate whether we are more interested in inference or prediction. Finally, provide $n$ and $p$.
a) We collect a set of data on the top 500 firms in the US. For each firm, we record record profits, number of employees, industry, and the CEO salary. We are interested in understanding which factors affect CEO salary.
b) We are considering launching a new product and wish to know whether it will be a _success_ or _failure_. We collect data on 20 similar products that were previously launched. For each product, we have recorded whether it was a success or failure, price charged for the product, marketing budget, competition price, and ten other variables.
c) We are interested in predicting the % change in the USD/Euro exchange rate in relation to the weekly changes in the world stock markets. Hence, we collect weekly data for all of 2012. For each week, we record the % change in the USD/Euro, the % change in the US market, the % change in the British market, and the % change in the German market.

### Solution
a) This is a regression problem because CEO salary is a continuous quantity, and the interest is in inference, as indicated by the phrase "we are interested in understanding". $n$ = 500, $p$ = 3.
b) This is a classification problem because the response is a discrete quantity, _success_ or _failure_. The interest here is in prediction because we want to know what the outcome will be for this particular "new product" we're launching. $n$ = 20, $p$ = 13.
c) This is a regression problem because % change is a continuous quantity. The problem tells us outright that it's one of prediction. $n$ = 52 (the number of weeks in a year), and $p$ = 3.

## Exercise 3
### Problem
We now revisit the bias-variance decomposition.
a) Provide a sketch of typical (squared) bias, variance, training error, test error, and Bayes (or irreducible) error curves, on a single plot, as we go from less flexible statistical learning methods towards more flexible approaches. The x-axis should represent the amount of flexibility in the method, and the y-axis should represent the values for each curve. There should be five curves. Make sure to label each one.
b) Explain why each of the five curves has the shape displayed in part (a).

### Solution
I'M GONNA SKIP THIS FOR NOW BECAUSE THE PLOTTING IS A PAIN

## Exercise 4
### Problem
You will now think of some real life applications for statistical learning.
a) Describe three real-life applications in which _classification_ might be useful. Describe the response, as well as the predictors. Is the goal of each application inference or prediction? Explain your answer.
b) Describe three real-life applications in which _regression_ might be useful. Describe the response, as well as the predictors. Is the goal of each application inference or prediction? Explain your answer.
c) Describe three real-life applications in which _cluster analysis_ might be useful.

### Solution
ai) We may want to understand (infer) if a student will pass or fail  a class on the basis of their sex, age, GPA, and major .
aii) We may want to predict if a political candidate will win or lose an election on the basis of their sex, age, political party, state, fundraising, and if they are the incumbent or not.
aiii) We may want to predict if a car will sell or not on the basis of make, model, year, and mileage.
bi) We may want to relate (infer) the mean delivery time of packets on a computer network on the basis of nodes, traffic over time, and an assortment of measures of network structure.
bii) We may want to predict the change in sales on the basis of radio, TV, and newspaper advertising budgets and location data.
biii) We may want to predict body fat as a function of neck, waist, and hip measurements, weight, height, and biological sex.
ci) We want to identify online shoppers who are likely to buy a certain product by finding those who have bought many items with the same tags/product descriptors as the one in question.
cii) Want to advertise a fuel rewards program by identifying drivers who drive many miles, use a diesel or gas engine, and what kind of vehicle they drive.
ciii) You want to identify residents of a neighborhood who are using many resources, such as water, electricity, and gas.

## Exercise 5
### Problem
What are the advantages and disadvantages of a very flexible (versus less flexible) approach for regression or classification? Under what circumstances might a more flexible approach be preferred to a less flexible approach? When might a less flexible approach be preferred?

### Solution
Focusing on regression, flexible methods are capable of producing more accurate predictions than less flexible methods. They perform well when we have a large sample size, when the variance of the error terms is small, and when the number of predictors is much smaller than the sample size. If any of these assumptions are violated, then a less flexible method will likely exhibit better performance.

## Exercise 6
### Problem
Describe the differences between a parametric and non-parametric statistical learning approach. What are the advantages of a parametric approach to regression of classification (as opposed to a non-parametric approach)? What are its disadvantages?



## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
